---
title: "Tidy Tuesday - BBC Best Hip Hop Song Rankings"
author: "Catalina Moreno"
date: "April 14, 2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = NA)
```

## Load libs

```{r}
library(tidyverse)
```

## Read in data

```{r}
polls <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-04-14/polls.csv')

rankings <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-04-14/rankings.csv')
```

## Initial eda 

### `polls` data


```{r}
polls %>% glimpse()
```

Critic info - check number of entries per unique critic:

```{r}
polls %>% count(critic_name, critic_rols, critic_country, critic_country2) %>% 
  count(n)
```

Looks good. Visualize `critic_country`:

```{r}
polls %>% 
  group_by(critic_country) %>% mutate(count = n()) %>% 
  ggplot(mapping = aes(x = fct_reorder(critic_country, count)))  + 
  geom_bar() + 
  coord_flip() +
  labs(x = "critic_country")
```

Critics are primarily from the US. 

```{r}
polls %>% count(critic_rols) %>% arrange(desc(n)) %>% head(n = 10)
```

A mix of roles and actual institution critic works for. 

Look at song info:

```{r}
polls %>% 
  count(title) %>% arrange(desc(n)) %>% head(n = 10)
```

Check all rankings over time:

```{r}
polls %>% 
  ggplot(mapping = aes(x = year, y = rank))+ geom_count()
```

Most ranked songs are in the early-mid 1990's. 


### `rankings` data 

Reproduce the ranking system:

```{r}
rankings_manual <- polls %>% mutate(score = 12 + -2*rank) %>% 
  group_by(title, artist) %>% summarise(total_points = sum(score),
                                total_votes = n(),
                                total_v1 = sum(rank == 1),
                                total_v2 = sum(rank == 2),
                                total_v3 = sum(rank == 3),
                                total_v4 = sum(rank == 4),
                                total_v5 = sum(rank == 5),
                                check = total_v1 + total_v2 + total_v3 + total_v4 + total_v5) %>% 
  # filter(check != total_votes)
  ungroup() %>% 
  arrange(desc(total_points),
          desc(total_votes),
          desc(total_v1), 
          desc(total_v2),
          desc(total_v3),
          desc(total_v4),
          desc(total_v5)) %>% 
  mutate(score = 1:n()) 
  
```

Join with `rankings` and see if reproduced results:

```{r}
rankings %>% 
  left_join(rankings_manual, by = c("ID" = "score")) %>% 
  filter(title.x != title.y) %>% nrow()
```

Many rows do not align. Check a few of the top entries for equivalency in ranking metrics (points, votes etc):

```{r}
rankings %>% head(n=20) %>% 
  left_join(rankings_manual %>% head(n = 20), by = c("ID" = "score")) %>% 
  filter(title.x != title.y) %>% nrow()
```

First 20 titles/scores match. Check if calcualted metrics match:

```{r}
rankings %>% head(n=20) %>% mutate(type = "original") %>% 
  select(-ID, -year, -gender) %>% 
  bind_rows(rankings_manual %>% head(n=20) %>% select(-score, -check) %>% 
              rename(points = total_points, 
                     n = total_votes,
                     n1 = total_v1,
                     n2 = total_v2, 
                     n3 = total_v3, 
                     n4 = total_v4,
                     n5 = total_v5) %>%   mutate(type = "manual")) %>% 
  gather(key = "key", value = "value", -artist,-title, -type) %>% 
  spread(key = "type", value = "value") %>% 
  ggplot(mapping = aes(x = manual, y = original)) +
  geom_point() + 
  geom_abline() +
  facet_wrap(~key, scales = "free")
```

An exact match. 

```{r}
rankings %>% anti_join(rankings_manual, by = c("title"))
```

And all titles overlap. 

Compare distinct `title`, `artist` pairs in `rankings` as it has more rows than `rankings_manual`:

```{r}
rankings %>% count(title, artist) %>% filter(n > 1)
rankings %>% filter(title %in% c("All Of The Lights"))
```

It lookes like "All Of The Lights" is incorrectly entered 2x, which means it's points are split among two entries. Check its existence in `polls` data:

```{r}
polls %>% filter(title %in% c("All Of The Lights"))
```

It appears that one critic entered `gender = male` when it should have been `mixed`. 

Solution: use `rankings_manual` in place of `rankings` to accurately represent "All Of The Lights" as one song. 

### Data correction

Recreate `rankings_manual` to mimic `rankings` more closely, first fixing `polls` incorrect entry:

```{r}
polls <- polls %>% mutate(gender = ifelse(title %in% "All Of The Lights", "mixed", gender))
```

```{r}
rankings_manual_final <- polls %>% 
  mutate(score = 12 + -2*rank) %>% 
  group_by(title, artist) %>% summarise(gender = unique(gender),
                                        year = unique(year),
                                        points = sum(score),
                                        n = n(),
                                        n1 = sum(rank == 1),
                                        n2 = sum(rank == 2),
                                        n3 = sum(rank == 3),
                                        n4 = sum(rank == 4),
                                        n5 = sum(rank == 5)) %>% 
  ungroup() %>% 
  arrange(desc(points),
          desc(n),
          desc(n1), 
          desc(n2),
          desc(n3),
          desc(n4),
          desc(n5)) %>% 
  mutate(ID = 1:n()) 
```

## Top songs eda

Look at top 20 ranked songs:

```{r}
rankings_manual_final %>% 
  slice(1:20) %>% 
  ggplot(mapping = aes(x = fct_reorder(title, points), y = points, fill = gender)) + 
  geom_col() + 
  coord_flip() +
  labs(x = "title", title = "Top 20 Ranked Hip-Hop Songs") +
  theme(legend.position = "bottom")
```

Evaluate all rankings by gender:

```{r}
rankings_manual_final %>% 
  ggplot(mapping = aes(x = year, y = points, color = gender)) + geom_point() +
  facet_wrap(~gender)

rankings %>% 
  ggplot(mapping = aes(x = gender, y = log(points))) + geom_boxplot(varwidth = T) + 
  stat_summary(fun.y = "mean", geom = "point", color = "red") +
  stat_summary(fun.y = "mean", geom = "line", color = "red", mapping = aes(group = 1))
```

Songs are mostly male, but mean points by gender shows males also score higher on average compared to females or mixed (note on log scale). 

"Juicy" won as the top ranked song by a landslide, but check for artists ranked across > 1 song:

```{r}
rankings_manual_final %>% 
  distinct(title, artist) %>% count(artist) %>% arrange(desc(n)) %>% 
  filter(n > 1) %>% 
  ggplot(mapping = aes(x = fct_reorder(artist, n), y = n)) + 
  geom_col() +
  coord_flip() +
  labs(x = "Artists w/ > 1 title ranked")
```

Jay-z's songs are ranked the most often across critics. How do the `points` for Jay'Z's songs compare to the leading song ("Juicy" by The Notorious B.I.G)?

```{r}
rankings_manual_final %>% 
  filter(str_detect(artist, "JAY")) %>% 
  ggplot(mapping = aes(x = fct_reorder(title,points), y = points, fill = artist)) + 
  geom_col() + 
  coord_flip() +
  geom_hline(yintercept = max(rankings_manual_final$points), color = "red", linetype = "dashed") +
  labs(x = "title", title = "Jay-Z's mentioned songs and scores") +
  theme(legend.position = "bottom")
```

Although Jay-Z has the most mentioned songs, their points are much lower than the top-ranked song. Also, we see Jay-Z has even more songs than counted above, as he appears as a solo artist as well as in collaborations with other artists. 


## Summary

This BBC dataset for ranking the top hip hop songs was very tidy. The only issue I caught was different `gender` entries for the song "All Of The Lights" which I believe should be considered `gender = mixed`. Because it had multiple entries (`mixed` and `male`) it was ranked in two separate entries of the `rankings` data set and did not get the correct `points` assignment. This means the `rankings` data set is incorrect starting at ranking 24: 



```{r}
rankings %>% 
  arrange(desc(points)) %>% 
  slice(1:30) %>% 
  mutate(type = "original") %>% 
  bind_rows(rankings_manual_final %>% 
              arrange(desc(points)) %>% 
              slice(1:30) %>%
              mutate(type = "corrected")) %>% 
  ggplot(mapping = aes(x = ID, y = type, color = ID > 23)) + 
  geom_text(mapping = aes(label = title), size = 3) +
  coord_flip() +
  scale_color_manual(values = c("TRUE" = "brown3", "FALSE" = "dodgerblue")) +
  labs(color = "Rankings Change After Correction", x = "Ranking", y = "Ranking Data Source",
       title = 'Effect of incorrect `gender` entry for "All Of The Lights" starts at song 24') +
  theme(legend.position = "bottom")


```

"Juicy" by The Notorious B.I.G is still by and far the leading ranked hip hop song, even after the data correction. 

