---
title: "Tidy Tuesday 05-21-2019"
subtitle: "Plastic Waste"
author: "Catalina Moreno"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      comment = NA)
library(tidyverse)
```

## Read in

```{r, readin}
coast_vs_waste <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/coastal-population-vs-mismanaged-plastic.csv")

mismanaged_vs_gdp <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/per-capita-mismanaged-plastic-waste-vs-gdp-per-capita.csv")

waste_vs_gdp <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/per-capita-plastic-waste-vs-gdp-per-capita.csv")

```

### coast_vs_waste

```{r, eda1}
coast_vs_waste %>% head()

coast_vs_waste %>% purrr::map_df(function(x) {sum(is.na(x))}) %>% 
  knitr::kable(caption = "Count of Missing Values")
```

Appears to be a lot of missing data. 

```{r, eda1b}
coast_vs_waste %>% na.omit() %>% nrow() 
coast_vs_waste %>% na.omit() %>% distinct(Year)
# coast_vs_waste %>% na.omit() %>% distinct(Entity) 
coast_vs_waste %>% na.omit() %>% count(Entity) %>% count(n)
```

Full data for year 2010, 175 different countries. 

```{r}
coast_vs_waste %>% na.omit() %>% ggplot(mapping = aes(x = `Mismanaged plastic waste (tonnes)`)) + geom_histogram() + facet_wrap(~Year)
```

Right-skew, try log-transform:

```{r}
coast_vs_waste %>% na.omit() %>% ggplot(mapping = aes(x = log(`Mismanaged plastic waste (tonnes)`))) + geom_histogram() + facet_wrap(~Year)

```

Check relationship between Population and Mismanaged Plastic Waste:


```{r}
coast_vs_waste %>% 
  select(`Mismanaged plastic waste (tonnes)`, `Total population (Gapminder)`, `Coastal population`) %>% 
  tidyr::gather(key = "key", value = "value", -`Mismanaged plastic waste (tonnes)`) %>% 
  ggplot(mapping = aes(x = log(value), y = log(`Mismanaged plastic waste (tonnes)`), color = key)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  facet_wrap(~key) +
  theme(legend.position = "bottom") +
  labs(x = "log(population)", title = "Log-Log Relationship")
```

Strong relationship between log-pop and log-mismanaged plastic. 

Check pop vs time:
```{r, pop_vs_time}
coast_vs_waste %>% 
  ggplot(mapping = aes(x = Year, y = `Total population (Gapminder)`, group = Code)) + geom_line()
```


Which are the two lines with largest populations? 

```{r}
coast_vs_waste %>% filter(`Total population (Gapminder)` > 5e8) %>% distinct(Code)
```

Can bring in continent info and color?

```{r}
library(countrycode)
continents <- codelist %>%
  select(iso3c, continent)

coast_vs_waste %>% 
  left_join(continents %>% select(Code = iso3c, continent), by = "Code") %>% 
    filter(!is.na(continent)) %>% 
  ggplot(mapping = aes(x = Year, y = `Total population (Gapminder)`, group = Code, color = continent)) + geom_line() + facet_wrap(~continent)
```

### mismanaged_vs_gdp

```{r, eda2}
mismanaged_vs_gdp %>% head()
```

Assume Total pop is same as coast_vs_waste, check:

```{r}
## check join
coast_vs_waste %>% select(Code, Year, `Total population (Gapminder)`) %>% 
  inner_join(mismanaged_vs_gdp %>% select(Code, Year, `Total population (Gapminder)`),
             by = c("Code", "Year")) %>% 
  ggplot(mapping = aes(x = `Total population (Gapminder).x`, y = `Total population (Gapminder).y`)) +
  geom_point(alpha = 0.5) + 
  geom_abline()
```

Same. Check missing values in this dataset:

```{r}
mismanaged_vs_gdp %>% purrr::map_df(function(x) {sum(is.na(x))}) %>% knitr::kable(caption = "Count of Missing Values")

mismanaged_vs_gdp %>% na.omit() %>% distinct(Year)
```

Also 2010 only. 

Check GDP per capita over time:
```{r}
mismanaged_vs_gdp %>% filter(!is.na(`GDP per capita, PPP (constant 2011 international $) (Rate)`)) %>% 
  ggplot(mapping = aes(x = Year, y = `GDP per capita, PPP (constant 2011 international $) (Rate)`, group = Code)) + geom_line()
```

One group that looks like >1 value per year?
```{r}
mismanaged_vs_gdp %>% 
  count(Entity, Year) %>% 
  filter(n > 1) 
```


Nothing comes up. 

Visualize per capita GDP vs per capita mismanaged waste, by continent:

```{r}
mismanaged_vs_gdp %>% filter(!is.na(`GDP per capita, PPP (constant 2011 international $) (Rate)`)) %>% 
  left_join(continents %>% select(Code = iso3c, continent), by = "Code") %>% 
  filter(!is.na(continent)) %>% 
  ggplot(mapping = aes(x = log(`GDP per capita, PPP (constant 2011 international $) (Rate)`),
                       y = log(`Per capita mismanaged plastic waste (kilograms per person per day)`),
                       color = continent)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~continent) + 
  labs(y = "log(kg mismanged plastic/person-day)", x = "log(GDP per capita)") +
  theme(legend.position = "bottom")
```

All continents except Africa show decrease in mismanaged plastic with increase in GDP in 2010.

### waste_vs_gdp


```{r, eda3}
waste_vs_gdp %>% head()
waste_vs_gdp %>% na.omit() %>% distinct(Year)
```

2010 only as well. 

Lastly compare per capita plastic vs per capit mismangaged plastic, bring in continent information, too:

```{r}
mismanaged_vs_gdp %>% select(mismanaged = `Per capita mismanaged plastic waste (kilograms per person per day)`, Code, Year) %>% 
  filter(!is.na(mismanaged)) %>% 
  inner_join(waste_vs_gdp %>% 
               select(Code, Year, plastic= `Per capita plastic waste (kilograms per person per day)`), 
             by = c("Code", "Year")) %>% 
  left_join(continents %>% select(continent, Code = iso3c), by= "Code") %>% 
  ggplot(mapping = aes(x = plastic, y = mismanaged, color = continent)) + 
  geom_point() + 
  facet_wrap(~continent, scales = "free_x") + 
  theme(legend.position = "bottom")
```

