---
title: "TidyTuesday 01-26-2021"
author: "Catalina Moreno"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE, 
                      warning = FALSE,
                      comment = NA)
```

## Background

This data comes from the Break Free From Plastic (BFFP) initiative (via TidyTuesday) and contains counts of plastic waste from a global audit conducted by volunteers. The plastic waste is categorized per recycling codes and the company that produced the waste is provided. This specific data from BFFP spans 2019-2020.

## Read in data

```{r}
## load libs
library(tidyverse)
```

Point to TidyTuesday GitHub to access data:

```{r}
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-01-26/plastics.csv')

plastics %>% head() %>% knitr::kable()
```

## Check the data

### uniqueness

Check if each country, year, parent_company entry is unique:

```{r}
plastics %>% count(country, year, parent_company) %>% filter(n > 1)
```

Looks good. Assign unique identifier to support any further data cleaning. 

```{r}
plastics <- plastics %>% mutate(uid = 1:n()) %>% 
  select(uid, everything())
```


### confirm `grand_total`

As each row is unique, group by `country`, `year`, `parent_company` and check count across plastic rows is equivalent to `grand_total`. 


Side note, first time trying `tidyr::pivot_longer() `as replacement to `tidyr::gather()`!

```{r}
plastics %>% 
  tidyr::pivot_longer(!c(uid, country, year, parent_company, 
                         grand_total, num_events, volunteers)) %>% 
  group_by(uid, country, year, parent_company) %>% 
  summarise(grand_total = first(grand_total),
            total_check = sum(value)) %>% 
  filter(grand_total != total_check)
```

Not bad, two observations with data entry issue, drop these. 

```{r}
mis_entry <- plastics %>% 
  tidyr::pivot_longer(!c(uid, country, year, parent_company, 
                         grand_total, num_events, volunteers)) %>% 
  group_by(uid) %>% 
  summarise(grand_total = first(grand_total),
            total_check = sum(value)) %>% 
  filter(grand_total != total_check) %>%
  pull(uid)

plastics <- plastics %>% filter(!uid %in% mis_entry)
```

### check zero or NA grand_total

Confirm each entry has at least one count (in `grand_total`):

```{r}
plastics %>% filter(is.na(grand_total) | grand_total == 0) %>% nrow()
```


Update data set to only have `grand_total > 0`:

```{r}
plastics <- plastics %>% filter(grand_total > 0)
```

### parent_company

As this is volunteer-submitted data that includes multiple countries/languages, differences in how `parent_company` is entered is likely. Convert to all caps as first step to standardize, though more text processing would be useful. 

```{r}
plastics <- plastics %>% mutate(parent_company = str_to_upper(parent_company))
```

A first skim of data set top counts, shows `NULL` and `UNBRANDED` and `GRAND TOTAL`

```{r}
plastics %>% select(parent_company, grand_total) %>% 
  arrange(desc(grand_total)) %>% 
  head(n = 10)
```

I (naively checked if `GRAND TOTAL` and `UNBRANDED` are companies...) but cross-referencing the great Shiny app published on the BFFP page (scroll down) [here](https://www.breakfreefromplastic.org/globalbrandauditreport2020/) for the US data entry in 2020:

```{r}
## can reproduce shiny?
plastics %>% 
  filter(country %in% 'United States of America', year %in% '2020') %>%
  arrange(desc(grand_total)) %>% 
  select(parent_company, year, grand_total) %>% 
  head()
```

As `UNBRANDED` and `NULL` are not in this app, assume these are invalid entries for `parent_company`. 

```{r}
# plastics <- plastics %>% filter(!parent_company %in% c("UNBRANDED", "NULL", "GRAND TOTAL"))
```

### country

Also capitalize country:
```{r}
plastics <- plastics %>% mutate(country = str_to_upper(country))
```

## Exploratory data analysis

### country

```{r}
plastics %>% distinct(country) %>% nrow()
```


Visualize top 10:

```{r}
plastics %>% 
  count(country) %>% arrange(desc(n)) %>% 
  head(n =10)
```

`country == EMPTY` highlights another flag with data entry. 

### plastic types 

```{r}
plastics %>% 
  tidyr::pivot_longer(!c(uid, country, year, parent_company, 
                         grand_total, num_events, volunteers)) %>% 
  group_by(year, name) %>% 
  summarise(total_count_per_plastic_group = sum(value, na.rm = TRUE)) %>% 
  ggplot(mapping = aes(x = forcats::fct_reorder(name, total_count_per_plastic_group), 
                       y = total_count_per_plastic_group)) + 
  geom_col(fill = "dodgerblue") + coord_flip() + facet_wrap(~year) +
  labs(x = "plastic type")
```

What are the implications of categorizing plastic as other?


### year

2019 to 2020 plastic count due to less plastic / COVID, or due to less volunteers?

```{r}
plastics %>% 
  ggplot(mapping = aes(x = year)) + geom_bar(fill = "dodgerblue") +
  labs(title = "# of submissions to BFFP by year")
```

```{r}
## # pieces of plastic
plastics %>% 
  group_by(year) %>% summarise(year_grand_total = sum(grand_total)) %>% 
  ggplot(mapping = aes(x = year, y = year_grand_total)) + geom_col(fill = "dodgerblue") +
  labs(title = "# of pieces of plastic by year")

## # of events
plastics %>% 
  group_by(year) %>% summarise(year_total_events = sum(num_events)) %>% 
  ggplot(mapping = aes(x = year, y = year_total_events)) + geom_col(fill = "dodgerblue") +
  labs(title = "# of counting events by year")
```



### parent_company

```{r}
plastics %>% 
  count(parent_company) %>% arrange(desc(n)) %>% 
  head(n = 10)
```
text processing needed for parent company, for example:

```{r}
plastics %>% 
  filter(str_detect(parent_company, "PHILIP MORRIS")) %>%
  distinct(parent_company)
```


### data collection

Since each row represents the count pertaining to a `parent_company` it does not capture a specific count event. Data is entered yearly. 

Which country has the most volunteers combined across 2019 and 2020?


```{r}
plastics %>% 
  count(country, volunteers, num_events) %>% 
  arrange(desc(volunteers)) %>% 
  head(n = 10) %>% 
  # arrange(desc(volunteers))
  ggplot(mapping = aes(x = forcats::fct_reorder(country, volunteers),
                       y = volunteers)) +
  geom_col(fill = "dodgerblue") + coord_flip() + 
  labs(x = "country", title = "Countries with the most volunteers 2019-2020")
```

Nice effort across these countries!

Next visualize the relationship between # of volunteers and # of documented pieces of plastic waste:

```{r}
plastics %>% 
  group_by(country, year) %>% 
  summarise(num_volunteers = unique(volunteers),
            grand_total_across_events = sum(grand_total)) %>% 
  ggplot(mapping = aes(x = num_volunteers, y = grand_total_across_events)) + 
  geom_point() +
  scale_y_log10() + scale_x_log10() + geom_smooth(method = "lm") +
  facet_wrap(~year) + labs("#counted plastic waste vs. # volunteers")
```

As expected, there is a positive relationship between the number of volunteers and the number of counted pieces of plastic waste. Call for more volunteers!

### source of plastic

Identifying top contributors to plastic waste is important to highlighting order to understand where our waste is coming from and try to improve:



```{r, fig.width = 10}
plastics %>% 
  group_by(parent_company) %>% summarise(total_count = sum(grand_total)) %>% 
  arrange(desc(total_count)) %>% 
  filter(total_count > 2000) %>% 
  mutate(parent_company = str_sub(parent_company, 1, 50)) %>% # cutoff length for presentation purposes
  filter(!parent_company %in% c("GRAND TOTAL", "NULL", "UNBRANDED")) %>% 
  ggplot(mapping = aes(x = forcats::fct_reorder(parent_company, total_count),
                       y = total_count)) + geom_col(fill = "dodgerblue") + coord_flip() +
  labs(x = "parent_company", title = "Top companies recorded that contribute to waste") 
```


As mentioned earlier, text processing of companies to clean up and categorize into high-level groups could help narrow in on biggest source / where to focus efforts. 

For an initial check, try categorizing by water, soda (coke and pepsi), coffee (starbucks), cigarettes (philip moris), candy (mars):

```{r}
plastics %>% 
  mutate(parent_company = str_to_upper(parent_company)) %>% 
  mutate(type = ifelse(str_detect(parent_company, "WATER"), "water", NA),
         type = ifelse(str_detect(parent_company, "PEPSI|COCA-COLA"), "soda", type),
         type = ifelse(str_detect(parent_company, "PHILIP MORRIS"), "cigarettes", type),
         type = ifelse(str_detect(parent_company, "MARS"), "candy", type),
         type = ifelse(str_detect(parent_company, "STARBUCKS"), "coffee", type)) %>% 
  filter(!is.na(type)) %>% 
  group_by(type) %>% summarise(total_count = sum(grand_total)) %>% 
  arrange(desc(total_count)) %>% 
  ggplot(mapping = aes(x = forcats::fct_reorder(type, total_count),
                       y = total_count)) + geom_col(fill = "dodgerblue") + coord_flip() + 
  labs(x = "type", title = "Initial try at categorizing waste")

```

Although categorizing could help focus campaigns, it doesn't show responsible companies (which I view as a limitation). 

## Summary

This data set helps bring awareness to plastic pollution per BFFP initiative. Very cool that volunteers contribute the data! Looking forward to monitoring the [BFFP website](https://www.breakfreefromplastic.org/globalbrandauditreport2020/)
